The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-08 rational.tex
intro:

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type $a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

Now consider the following algebraic manipulation (assuming $a \neq 0$):
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
\end{array}\]
As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
{{a <-> a}} isomorphism commputes as follows:
\begin{itemize}
\item We start at line (0) with the value {{20}};
\item We proceed to line (1) with the value {{(20,())}};
\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
  $\alpha$ is essentially a logic variable;
\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
  forces the unification $\alpha=20$, i.e., the value becomes
  {{((20,1/20),20)}} which simplifies to {{((),20)}};
\item We reach line (5) with the value {{20}}.
\end{itemize}

\amr{the above is a boring trace; we need to convey why it is exciting but we
  are not agreeing on the interpretation.}

@@

The reductions for the primitive isomorphisms are exactly the same as
have been presented before \cite{infeffects}. The reductions for the
closure combinators are however presented in a small-step operational
style. The following are the definition for evaluation contexts and
machine states:

@@

The reduction rules below essentially track the flow of particles
through their circuit. The start machine state, {{<c,v,[]>}}, denotes
the particle/value {{v}} about to be evaluated by the circuit
{{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
where the particle {{v}} has exited the circuit {{c}}.

%% For the non-primitive combinators, we express the semantics using evaluation
%% contexts which are used to represent the ``rest of the computation.'' More
%% precisely, to calculate the result of applying combinator $c$ to value $v$,
%% we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
%% transitions of the machine (see below) are unusual in the sense that they
%% always maintain full information about {{c}}, although this information may
%% be split in the first and last component of the machine. The {{v}} component
%% of the machine is the one that captures the current value as it is flowing in
%% the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
%% transitions based on the structure of {{c}}. When the evaluation of {{c}} is
%% complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
%% restored.

%% This is a mess. Perhaps give a small example and explain things based on the
%% example before giving the full transitions.

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, C> &|-->& {[iso, v', C]} & (1)
% & & where iso v |--> v' &
% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (2) 
% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (3) 
% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (4) 
% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (5) 
% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (6)
% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (7)
% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (8)
% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (9)
% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (10) 
% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (11)

The rule (1) describes evaluation by a primitive isomorphism. Rules
(2), (3) and (4) deal with sequential evaluation. Rule (2) says that
for the value {{v}} to flow through the sequence {{c1 (;) c2}}, it
should first flow through {{c1}} with {{c2}} pending in the context
({{Fst C c2}}). Rule (3) says the value {{v}} that exits from {{c1}}
should proceed to flow through {{c2}}. Rule (4) says that if the value
{{v}} has exited {{c2}} as well, then it is equivalent to having
exited from {{c1(;)c2}}.

Rules (5) to (8) deal with {{c1 (+) c2}} in the same way. In the case
of sums, the shape of the value,i.e. is it is tagged with {{left}} or
{{right}}, determines whether path {{c1}} or path {{c2}} is
taken. Rules (9), (10) and (11) deal with {{c1 (*) c2}} similarly. In
the case of products the value should have the form {{(v1, v2)}} where
{{v1}} flows through {{c1}} and {{v2}} flows through {{c2}}. Both
these paths are entirely independent of each other and we could
evaluate either first, or evaluate both in parallel. In this
presentation we have chosen to follow {{c1}} first, but this choice is
entirely arbitrary.

The interesting thing about the semantics is that it represents a
reversible abstract machine. In other words, we can compute the start
state from the stop state by changing the reductions `{{|-->}}' to run
backwards `{{<--|}}'. We will take advantage of this property in
Sec. \ref{sec:rat}.

@@

\roshan{We really have to rethink this prop in the presence of
  negatives and fractionals. With negatives, the program can actually
  end at the beginning of the circuit -- i.e. the program {{c:b1<->b2}}
  can stop with a value of type {{b1}}. With fractionals, there can be
  several values of type {{b2}} produced for a value of type {{b1}}. }

@@

\roshan{I chose to draw the figure with b instead of a, since b is what
  we use for the types. Should this be a? Also I chose to make the
  swap and 0 wires, explicit.Is this what you intended? }

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

start with int times bool divided by bool: it is like an int but it can only
be used as the second component of a pair whose first component is an int;
gives you the power to say here is 20 but you must use it to buy books

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~$a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

% Now consider the following algebraic manipulation (assuming $a \neq 0$):
% \[\begin{array}{rclll}
% && a & \mbox{initial~$a$} & (0) \\
% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
% \end{array}\]
% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
% {{a <-> a}} isomorphism commputes as follows:
% \begin{itemize}
% \item We start at line (0) with the value {{20}};
% \item We proceed to line (1) with the value {{(20,())}};
% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
%   $\alpha$ is essentially a logic variable;
% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
%   forces the unification $\alpha=20$, i.e., the value becomes
%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
% \item We reach line (5) with the value {{20}}.
% \end{itemize}

% \amr{the above is a boring trace; we need to convey why it is exciting but we
%   are not agreeing on the interpretation.}

@@

For the non-primitive combinators, we express the semantics using evaluation
contexts which are used to represent the ``rest of the computation.'' More
precisely, to calculate the result of applying combinator $c$ to value $v$,
we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
transitions of the machine (see below) are unusual in the sense that they
always maintain full information about {{c}}, although this information may
be split in the first and last component of the machine. The {{v}} component
of the machine is the one that captures the current value as it is flowing in
the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
transitions based on the structure of {{c}}. When the evaluation of {{c}} is
complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
restored.

This is a mess. Perhaps give a small example and explain things based on the
example before giving the full transitions.

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, C> &|-->& {[iso, iso(v), C]}
% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2>
% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C>
% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]}
% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2>
% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]}
% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C>
% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]}
% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2>
% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C>
% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]}

@@

\roshan{We really have to rethink this prop in the precense of
  negatives and fractionals. With negatives, the program can actually
  end at the begining of the circuit -- i.e. the program {{c:b1<->b2}}
  can stop with a value of type {{b1}}. With fractionals, there can be
  several values of type {{b2}} produced for a value of type {{b1}}. }

@@

\roshan{I chose to draw the igure with b instead of a, since b is what
  we use for the types. Should this be a? Also I chose to make the
  swap and 0 wires, explicit.Is this what you intended? }
